---
title: "Class 11: Structural Bioinformatics"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## The PDB database for biomolecular structural data


Download a CSV file from the PDB site (“Analyze” -> “PDB Statistics” > “by Experimental Method and Molecular Type”)

> Q1: Determine the percentage of PFB structures solved by X-ray and Electron Microscopy.

```{r}
# Read CSV
data <- read.csv("Data Export Summary.csv")
data
```

Total number of entries
```{r}
sum(data$Total)
```


Proportion of entries from each method
```{r}
round((data$Total/sum(data$Total)) * 100, 2)
```
Vast majority of protein is done by X-ray: 89%
Electron Microscopy is only 2.49%


> Determine what proportion of structures are protein?

Proportion that are protein
```{r}
round (sum(data$Proteins) / sum(data$Total) * 100 , 2)
```


## HIV-Pr structure analysis

Here we will read the 1HSG PDB structure and select the protein component and write out a new **protein-only** PDB format file. We then do the same for the ligand (i.e. known drug molecule) creating a **ligand-only** PDB file.

```{r}
library(bio3d)
# R package: biological sutructure analysis
pdb<- read.pdb("1hsg.pdb")
pdb

#read.pdb()
#atom.select()
#write.pdb()
#trim.pdb()

```


```{r}
ligand <- atom.select(pdb, "ligand", value=TRUE)
write.pdb(ligand, file="1hsg_ligand.pdb")
# only give the 1hsg 里的 ligand 里的protein, only give you MK1
```





> Q2: Type HIV in the PDB website search box on the home page and determine how many HIV-1 protease structures are in the current PDB?

L ColorId 4 resid25
newcar chain protein
Pdw Name not protein and not water 








