---
title: "Class 7 R Functions and packages"
author: "Belinda Xue"
date: "10/22/2019"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# R Functions Revisited

Source my functions from last day

```{r}
source("http://tinyurl.com/rescale-R")
# function that run another file/source
# online file or can be local file
# the glasses on the tab name means "read only"
```

```{r}
rescale(1:10)
# take all numerical input and expel to 0-10
```

```{r}
rescale(c(1,10,5,NA,6))
# na.rm default is false
# but rescale function have na.rm= TRUE, so fix the NA already; otherwise will be all NA
```


```{r}
# is.numeric is asking whether those are numbers
is.numeric(1:10)
# asking wheter those are numbers
is.numeric(c(1,5,NA,10))
is.numeric(c(1,5,NA,"bb"))

# ! filp the orignal answers; from T to F
!is.numeric(c(1,5,NA,"bb"))
# orignal is false, but with ! would be true
!is.numeric(c(1,5,NA,10))
# orginail is true, but with ! would be False

```

the **!** 
```{r}
# rescale2(c(1:10, "bb")) give you error
# would said "Error: Input x should be numeric"
# rescale2 only want numbers!! 
```

## In-class example of writing function! both_na() example

Question is "How many TRUE values are there?"
Write a function to find where there are NA elements in two input vectors

First make some simple input where I know the answer
```{r}
# Lets define an example x and y
x <- c( 1, 2, NA, 3, NA) 
y <- c(NA,3,NA,3, 4)
```

Look online and found the **is.na()** function
And the **which()** fucntion tells me where the TRUE values are 
```{r}
is.na(x)
# give you either true or false for all vectors
# True thing is where NA is 
which(is.na(x))
# tell you exactly which element of vector is NA
# try with y too
is.na(y)
which(is.na(y))
```


The AND function requires two input TRUE to give a TRUE
```{r}
is.na(x) & is.na(y)
# the & symbol... only give you TRUE when both vector is TRUE; only input is NA will give you TRUE; only one NA? give you FALSE
```

Taking the **sum()** of TRUE FALSE vector will tell me how many TRUE elements I have. This is my working snippet!!
```{r}
sum(is.na(x) & is.na(y))
# sum up numbers of TRUE
sum(c(TRUE, TRUE, FALSE, TRUE))
```


### Now turn it into a function!

```{r}
both_na <- function(x, y) {
  # BODY is the statement/ working snippet
  sum(is.na(x) & is.na(y))
}
```

```{r}
both_na(x, y)
```
Test Function
```{r}
 x <-  c(NA, NA, NA)
y1 <- c(1, NA, NA)
y2 <- c(1, NA, NA, NA)
```

```{r}
both_na(x, y1)
# 有两个共同的TRUE; cuz got two column of NA
both_na(x, y2)
# will give you ?? cuz y2 is longer than x
# x is not as long as y2, will recycle x
```

```{r}
x2 <- c(NA,NA)
both_na(x2, y2)
```

Side note: This is re-cycle example of the colors
```{r}
plot(1:10, col=c("red","blue","green"))
# gonna recycle the color red and blue for each dot
# shorter vector will align with teh longer vector
```


```{r}
 x <-  c(NA, NA, NA)
y3 <- c(1, NA, NA, NA, NA, NA, NA)
both_na(x, y3)
# will make x the same length as y3
# view x as NA NA NA (NA NA NA NA)
# should get 6

x <-  c(1, NA, NA)
y3 <- c(1, NA, NA, NA, NA, NA, NA)
both_na(x, y3)
# will make x the same length as y3
# view x as 1 NA NA (1 NA NA 1)
# same as this x <- c(1, NA, NA, 1, NA, NA, 1)
# should get 4
```


```{r}

length(x)
length(y3)
# tell me how long is the x or y3
# match them? will get recylce problem

7 == 7
7 != 7
# does 7 equal to 7
# does 7 NOT equal to 7

```

```{r}
both_na2 <- function(x, y) {
  if(length(x) != length(y)) {
    stop("Inputs x and y should be the same length!")
    # the stop is where you input your error messange
  }
  sum(is.na(x) & is.na(y))
}
```

```{r}
# both_na2(x, y3)
# test # 
```








# Practice writing function! Grade example
```{r}
# Student 1
c(100, 100, 100, 100, 100, 100, 100, 90)
# Student 2
c(100, NA, 90, 90, 90, 90, 97, 80)
```

Name the vectors
```{r}
s1 <- c(100, 100, 100, 100, 100, 100, 100, 90)
s2 <- c(100, NA, 90, 90, 90, 90, 97, 80)
```

me testing some function
```{r}
drop_lowest <- sum(s1) - min(s1)
mean(s1)
```

Student 1 grades
```{r}
# which number in s1 is the lowest?
s1 [which.min(s1)]
# a - sign give you everthing except the min 90
s1 [ -which.min(s1)]
# take the mean of that => give you s1 avg after dropping the lowest grade
mean (s1 [ -which.min(s1)])
```

Student 2 Grades
**THIS DOES NOT WORK WHEN WE HAVE NA!! :(**
```{r}
# which number in s1 is the lowest?
s2 [which.min(s2)]
# a - sign give you everthing except the min 80
s2 [ -which.min(s2)]
# take the mean of that => give you s1 avg after dropping the lowest grade
mean (s2 [ -which.min(s2)])
```

Need to fix the NA for student 2!
```{r}

# give you the total length??
which.min(s2)

# which number in s1 is the lowest?
s2 [which.min(s2)]

# a - sign give you everthing except the min 80
s2 [ -which.min(s2)]

# take the mean of that => give you s1 avg after dropping the lowest grade
# ignore the NA by na.rm = TRUE
mean (s2 [ -which.min(s2)], na.rm= TRUE)
```

```{r}
# check whether s1 or s1 have NA using "ANY"
any(is.na(s1))
any(is.na(s2)) 

```


#### We have our working code! Now turn it into a first function! 

```{r}
grade <- function(x) {
  if( any(is.na(x)) ) {
    warning("Student is missing a homework!")
  }
  mean (x [ -which.min(x)], na.rm= TRUE)
  }
```

```{r}
grade(s1)
grade(s2)
```


###### ???? another example?? 
```{r}
grade2 <- function(x) {
  if( any(is.na(x)) ) {
    warning("Student is missing a homework!")
  }
  sum (x [ -which.min(x)], na.rm= TRUE) / (length(x)-1)
}
```

```{r}
s3 <- c(100, NA, NA, NA, NA)
grade2(s3)
```


#### side note of the grade example... 
```{r}
url <- "https://tinyurl.com/gradeinput"
hw <- read.csv(url, row.names=1)
```

```{r}
apply(hw, 1, grade)
# apply (x, margin, function, ...)
```


# Collabrative CRAN packages

```{r}
# Go to class website and find Lec 7 "Collaborative Google Doc based notes on selected R packages", click the link inside the doc
# our group is doing ggplot2 package
# install using "install.packages("tidyverse")"
# read the package using "library(ggplot2)"
library(ggplot2)
ggplot(mpg, aes(displ, hwy, colour = class)) + 
  geom_point()

# example of plotting the graph using ggplot2
```





```{r}
library(DESeq)
```











